---
title: "ROC Curve"
subtitle: "Stat 154, Fall 2017"
author: "Gaston Sanchez"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(AppliedPredictiveModeling)
library(caret)
library(ggplot2)
library(MASS)
library(pROC)
```

This document is based on the simulation study discussed in Chapter 11, from
_Applied Predictive Modeling_ by Kuhn and Johnson (2013).

The required R packages are:

```r
library(AppliedPredictiveModeling)
library(caret)
library(ggplot2)
library(MASS)
library(pROC)
```


## Introduction

"The ROC curve is used for determining alternate cutoffs for class probabilities.
This curve is created by evaluating the class probabilities for the model
across a continuum of thresholds. For each candidate threshold, the resulting
true positive rate (i.e. sensitivity) and the false-positive rate (1 - specificity)
are plotted against each other." (Kuhn and Johnson, 2013)"


-----


## Simulated Data

A simulated data set with two predictors, and one response variable with two
classes is obtained with `quadBoundaryFunc()`

```{r sim_train}
# simulated data with function quadBoundaryFunc() which
# creates a class boundary that is a function of both predictors. 
set.seed(975)
simulated_train <- quadBoundaryFunc(500)

head(simulated_train)
```

Let's see what the data looks like:

```{r scatter1, fig.align='center'}
ggplot(simulated_train, aes(x = X1, y = X2)) + 
  geom_vline(xintercept = 0, color = 'gray70') + 
  geom_hline(yintercept = 0, color = 'gray70') +
  geom_point(aes(color = class), alpha = 0.8)
```

We can add density curves to visualize the ellipses:

```{r scatter2, , fig.align='center'}
ggplot(simulated_train, aes(x = X1, y = X2)) + 
  geom_vline(xintercept = 0, color = 'gray70') + 
  geom_hline(yintercept = 0, color = 'gray70') +
  geom_density2d(color = 'gray60') +
  geom_point(aes(color = class), alpha = 0.8)
```


-----


## Linear Discriminant Analysis (LDA)

We begin the classification model building process by fitting a model using
linear discriminant analysis. This is done with the function `lda()` from 
`"MASS"`:

```{r lda}
# LDA
lda_fit <- lda(class ~ X1 + X2, data = simulated_train)
lda_pred_train <- predict(lda_fit, simulated_train)
```

Using the predictions, we can get the following confussion matrix:

```{r lda_confussion}
lda_confussion <- table(obs = simulated_train$class, pred = lda_pred_train$class)
lda_confussion
```

and the associated _default_ error rate given by:

```{r}
# error rate (misclasification rate)
1 - sum(diag(lda_confussion)) / nrow(simulated_train)
```


-----


### Quadratic Discriminant Analysis (QDA)

The second model involves fitting a QDA with the function `qda()` from 
`"MASS"`:

```{r qda)}
# QDA
qda_fit <- qda(class ~ X1 + X2, data = simulated_train)
qda_pred_train <- predict(qda_fit, simulated_train)
```

Confussion matrix:

```{r qda_confussion}
qda_confussion <- table(obs = simulated_train$class, pred = qda_pred_train$class)
qda_confussion
```

and the _default_ error rate is:

```{r qda_error}
1 - sum(diag(qda_confussion)) / nrow(simulated_train)
```


-----


## ROC Curve


Add LDA and QDA posterior probabilities for Class 1 to the data frame of 
simulated data:

```{r}
# adding predicted class-1 probabilities
simulated_train$ldaprob <- lda_pred_train$posterior[ ,"Class1"]
simulated_train$qdaprob <- qda_pred_train$posterior[ ,"Class1"]
```


One way to obtain ROC curves is with the `"pROC"` package. To get such a curve,
you need to create a `"roc"` object with the function `roc()`.

```{r lda_roc}
# ROC Curve for LDA
lda_train_roc <- roc(
  response = simulated_train$class,
  predictor = simulated_train$ldaprob,
  # this function assumed that the second class
  # is the event of interest, so labels msut be reversed
  levels = rev(levels(simulated_train$class))
)

# ROC curve
plot(lda_train_roc, legacy.axes = TRUE, las = 1)
```

The ROC curve for QDA can be obtained as follows:

```{r qda_roc}
# ROC Curve for QDA
qda_train_roc <- roc(
  response = simulated_train$class,
  predictor = simulated_train$qdaprob,
  levels = rev(levels(simulated_train$class))
)

# ROC curve
plot(qda_train_roc, legacy.axes = TRUE, las = 1)
```

Both curves, can be plotted in the same figure:

```{r lda_qda_curves}
# ROC curves
plot(lda_train_roc, legacy.axes = TRUE, las = 1, 
     col = "#2f74e2", lwd = 3)
plot(qda_train_roc, legacy.axes = TRUE, las = 1, 
     col = "tomato", lwd = 3, add = TRUE)
```
